---
title: Towards A Workflow For Exploring Missing Values
author: Timothy Deitz, Clinical Psychologist & Aspiring Data Scientist
date: '2019-09-18'
categories:
  - R
  - plotly
  - ggplot2
  - naniar
tags:
  - plotly
  - ggplot2
  - naniar
output:
  blogdown::html_page:
    toc: yes
---
<style>
body.blue { background-color:#2b3f60;}
</style>

<body class = "blue">


```{r setup, echo = TRUE, include=FALSE}

library(dplyr)
library(ggplot2)
library(tidyr)
library(broom)
library(forcats)
library(purrr)
library(plotly)
library(tibble)
library(naniar)
library(visdat)
library(DT)

```


##Setting Up

Missing values are often inevitable, and typically a source of anxiety amoung people responsible for analysing data. This post aims to sketch out the beginnings of a workflow for missing value exploration and identification. Thanks to [Nicholas Tierney ](https://github.com/njtierney) for the excellent `naniar` and `visdat` packages from which this post draws inspiration.

Let's begin by simulating some psychological data. I've deliberately made a small dataframe so that analysis steps are easier to understand.

Each character string in the `scale` column represents a psychological outcome measure. The PHQ9 is a depression measure, and the GAD7 is an anxiety measure. The data represent scores across these measures for different individuals at different points in time.

```{r}

data<- tibble::tribble(
  
  ~scale, ~date,        ~ score, ~pts,   ~id,                    ~sex,
  "PHQ9", "02/02/2003",   10,      9,     "",                     "m",  
  "GAD7","02/03/2003",   12,      11,    uuid::UUIDgenerate(),   "f",
  "PHQ9", "01/01/2004",  8,       9,     uuid::UUIDgenerate(),   999,
  "GAD7", "04/02/2005",  8,       ".",   uuid::UUIDgenerate(),   "m",
  "GAD7", "04/02/2005",  3,       4,     uuid::UUIDgenerate(),   "f", 
  "PHQ9", "04/02/2005",  4,       2,     "",                     -98,   
  "GAD7", "09/02/2006",  1,       ".",   uuid::UUIDgenerate(),   "m",
  "PHQ9", "02/02/2003",  1,       1,     uuid::UUIDgenerate(),   -999,
  "GAD7","02/03/2003",   9,       8,     uuid::UUIDgenerate(),   "f",
  "PHQ9", "N/A",         7,       8,     uuid::UUIDgenerate(),   999,
  "GAD7", "04/02/2005",  4,       7,     uuid::UUIDgenerate(),   "f",
  "GAD7", "04/02/2005",  2,       4,     "",                     "m",
  "PHQ9", "N/A",         8,       7,     uuid::UUIDgenerate(),   -98,
  "GAD7", "N/A",         8,       7,     uuid::UUIDgenerate(),   "f"
  
)

```

Here's what the data looks like

```{r echo = FALSE}

data

```

As you can tell, it is a disaster in terms of consistent coding of missing values in the dataframe. A whole bunch of diffirent values have seemingly been used to denote missingness.

Before doing much else, it's often a good idea to take a look at the range of unique values for each variable — this lets us identify odd values that may either be mistakes or missing value codes.

We can see unique values across all variables quickly as follows:

```{r echo = TRUE}

purrr::map(data, unique)

```

Here it's easy to see that missing values have been coded with -999, -98, ., "N/A" and "" - depending on the variable.

If we're reading the data in from a CSV, it's easy enough to specify these strings at the time of import, something like this:


Assuming our file is called outcomes.csv

```{r echo = TRUE, eval = FALSE}

data<- readr::read_csv("outcomes.csv", 
                       na = c("N/A", ".", "", "-999", "-98"))

```

Of course, readr will try to guess the column types automatically when importing.
If there are odd values in numeric columns, these will be read in as character variables and we would have to coerce them to numeric after the fact.

But sometimes, depending on how/where we get our data, we don't have the option of explicitly declaring missing value codes at import time.

This is where the naniar package becomes really useful. The code below will replace any of the specified codes (e.g. -999) with explicit NA values.

```{r echo = TRUE}

na_vals<- c("N/A", ".", "", "-999", "-98")

data_na<- data %>% naniar::replace_with_na_all(condition = ~.x %in% na_vals)

#View the data after applying the function

data_na

```

As you can see, missing values are now explicit and not represented by heterogeneous digits/letterss. 

Sometimes we only want to target a selection of variables, under certain conditions. This might be helpful if there are legitimate values of -999 or -98 in some columns. We could do this as follows: 

```{r echo = TRUE, eval = FALSE}

#Only code missing values for `sex` and `id`

data_na<- data%>% replace_with_na_at(.vars = c("sex", "id"), 
                                  condition = ~.x %in% na_vals)

```

or

```{r echo = TRUE, eval = FALSE}

#Only apply missing value coding to character variables

data_na<- data %>% naniar::replace_with_na_if(.predicate = is.character, 
                                              condition = ~.x %in% na_vals)

```

##Overviewing missingness at level of the dataframe

I think it can be helpful to have mental template of exploring missing values on three levels: (1) At the level of the whole dataframe, (2) at the level of rows (i.e. cases), and (3) at the level of individual variables.

Let's start with the whole dataframe. We can quickly find out how many missing values and complete values exist, as follows:

```{r echo = TRUE}

#Count missing values
naniar::n_miss(data_na)

#Count complete values 
naniar::n_complete(data_na)

```

And to calculate the percentage of missing values in the dataframe:

```{r echo = TRUE}

#Get proportion/percentage of missing values
n_miss(data_na) / n_complete(data_na)

```

##Row-Wise Missing Value Exploration

Now we move onto row level analysis. This is important because certain statistical analyses require a minimum degree of completeness for cases included.

Firstly, what percentage of rows have missing data?

```{r echo = TRUE}

pct_miss_case(data_na)

```

Let's get a row-by-row summary of missing data info:

```{r echo = TRUE}

#Save to object because we'll use this output again in a moment

case_missing_summary<- miss_case_summary(data_na)

#Let's view the summary

case_missing_summary

```

This output tells us, for example, that case 6 has two values missing, constituting 33% missing values within his/her row (i.e. across all variables fow which he or she has data).

But I think it is actually easier to visualise this information graphically, so let's do that:

```{r echo = TRUE}

#Filter to capture only cases with missing data
all_missing<- data_na[which(case_missing_summary$n_miss != 0),]

#Summarise these cases and add an identifier column (whichever you want, in thise case `id`)
all_missing_summary<- miss_case_summary(all_missing) %>% 
  dplyr::mutate(identifier = all_missing$id)

#Make the bar plot of missingness by row
missing_by_row_plot<- ggplot(all_missing_summary, 
                             aes(x = fct_reorder(identifier, pct_miss), 
                                 y = pct_miss)) + 
  geom_bar(stat = "identity") + coord_flip() + xlab("Case") + 
  ylab("Percentage Missing") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L, scale = 1)) 

#Add some plotly tooltip customisation to give extra info
missing_by_row_plot<- missing_by_row_plot + 
  aes(text = paste("N Missing: ", n_miss, "<br>",
                    "% Missing: ", round(pct_miss, 2), "<br>",
                    "Case Number:", case))
#Render the plot
ggplotly(missing_by_row_plot, tooltip = "text")

```

Another helpful way of evaluting row-wise missingness is using a raster or tile plot. The `visdat()` function from the visdat package is great for this, but we'll rotate it to have row identifiers on the y-axis.

```{r}

vis_dat_plot<- visdat::vis_dat(data_na) + coord_flip()

ggplotly(vis_dat_plot)

```

##Variable-Wise Missing Data Exploration

First, let's work out what percentage of our variables having missing values:

```{r echo = TRUE}

prop_miss_var(data_na)

```

Next, we want a break-down of how many values (and what percentage) are missing per variable:

```{r echo = TRUE}

missing_by_var<- miss_var_summary(data_na)

#View the breakdown

missing_by_var

```

But again, it helps to visualise this information (and don't forget to hover over the plot to get the extra
info we added when customising the tooltip).

```{r echo = TRUE}

missing_by_var_plot<- ggplot(missing_by_var, 
                             aes(x = fct_reorder(variable, pct_miss), 
                                 y = pct_miss)) + 
  geom_bar(stat = "identity") + coord_flip() + xlab("Variable") + 
  ylab("Percentage Missing") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L, scale = 1)) 

  missing_by_var_plot<- missing_by_var_plot + 
    aes(text = paste("N Missing: ", n_miss, "<br>",
                     "% Missing: ", round(pct_miss, 2)))
  
  ggplotly(missing_by_var_plot, tooltip = "text")
  
```

As an extra, if we want to break down missing value across variables **within levels** of a grouping variable, we can just do this:

```{r echo = TRUE}

data_na %>% group_by(scale) %>% miss_var_summary()

```

Hope this was helpful, any feedback welcome of course.

</body>
