---
title: "Supercharged Data Cleaning with R"
author: Timothy Deitz, Clinical Psychologist & Data Science Consultant
date: '2019-11-23'
categories:
  - Data Cleaning
tags:
  - Data Cleaning
output:
  blogdown::html_page:
    toc: yes
---
<style>
body.blue { background-color:#2b3f60;}
</style>

#Supercharged Data Cleaning with R — Part One

##Importing the Data

Real-world data is messy, irritating and cumbersome. Many tutorials on data wrangling with R simply don't prepare you for it. The following article will be the first in a series of expositions analysing unmanicured datasets using best practices. As I am a Clinical Psychologist by training, the content will be skewed towards mental health and psychology-related datasets.

I begin by loading all required packages.

```{r}
library(fs)
library(readr)
library(here)
library(purrr)
library(janitor)
library(dplyr)
library(stringr)
library(naniar)
library(plotly)
library(stringr)
library(forcats)

```

The `fs` package is fantastic for anything that relates to file/folder navigation using `R`. The function `here()` always finds the folder where the `.Rproj` file is contained, and treates this folder as the working directory. Because I am not specifying an absolute file path (e.g. `tim_data\analysis\import`), other people who have the data can run my script immediately.

Next I list all files in the working directory to locate the spreadsheets in which my data is contained.

```{r}
fs::dir_ls(path = here())

```


Because I only want csv files, I use a basic regular expression to import only files that end in `.csv`. Then, I use the `map()` to programatically read in and store each dataframe (i.e. from each spreadsheet) into a list of dataframes. I will later bind them into a single dataframe.

```{r}
csv_files<- fs::dir_ls(here(), regexp = "\\.csv$")

pain_data_raw<- csv_files %>% purrr::map( ~readr::read_csv(.x)) %>% 
  purrr::set_names("early_19", "mid_17", "mid_18")
```

The `readr` automatically guesses the data type of each imported column. But it doesn't always guess right, and this can create problems with large datasets. In theory, each of the three dataframes should have identical variables. But let's check this, using a handy function from the `janitor` package.

```{r}
data_comparison<- janitor::compare_df_cols(pain_data_raw)

data_comparison %>% dplyr::filter(column_name == 'EEnd_BPIPainInterference')

```

As you can see, the `EEnd_BPIPainInterference` column is specified as a numeric variable in two of the dataframes, but as a character varible in the other. To be clear, this variable (which measures how much pain disrupts daily activities) should certainly be numeric. The character specification in one of the dataframes is probably caused by an errant fullstop or other character in the column (a data entry error).

Therefore, when you have lots of columns (> 100) and likely data entry mistakes, it is often best to manually import all columns as character variables, and then alter their type as needed. I I can do this by adding `col_types = cols(.default = "c")` to the original code.

```{r}
pain_data_raw<- csv_files %>% purrr::map( ~readr::read_csv(.x, col_types = cols(.default = "c"))) %>% 
  purrr::set_names("early_19", "mid_17", "mid_18")

```

If I now compare each dataframe again, all variables are of type character. But there is another problem. Certain variables appear in one dataframe but not the others, or appear in two dataframes but not the third. This is indicated by `NA` values in the output.

```{r}
data_comparison

```

Since I want to analyse all the data, I can simply opt to **only** retain variables that are common to each dataframe. The first step here will be to store the names of those common variables in a character vector.

```{r}
matching_vars<- data_comparison %>% dplyr::filter_at(vars(-column_name), all_vars(!is.na(.))) %>% 
  dplyr::select(column_name) %>% dplyr::pull(1)

matching_vars
```


I then use `map()` once again, to iterate through each dataframe and select only these variables

```{r}

pain_data_complete<- pain_data_raw %>% purrr::map(~dplyr::select(.x, matching_vars))

```

To be sure it worked, I can check that each dataframe now has the same number of columns. 

```{r}
purrr::map_dbl(pain_data_complete, ncol)
```

Success!!!

Having succeeded at importing and merging the data, it is a good idea to store the above steps into a reusable function.

```{r}

import_data<- function() {

csv_files<- fs::dir_ls(here(), regexp = "\\.csv$")

pain_data_raw<- csv_files %>% purrr::map( ~readr::read_csv(.x, col_types = cols(.default = "c"))) %>% 
  purrr::set_names("early_19", "mid_17", "mid_18")

data_comparison<- janitor::compare_df_cols(pain_data_raw)

matching_vars<- data_comparison %>% dplyr::filter_at(vars(-column_name), all_vars(!is.na(.))) %>% 
  dplyr::select(column_name) %>% dplyr::pull(1)

pain_data_complete<- pain_data_raw %>% purrr::map(~dplyr::select(.x, matching_vars))

joined_data<- dplyr::bind_rows(pain_data_complete)

}

```

Test that it works

```{r}

joined_data<- import_data()

glimpse(joined_data)

```


<body class = "blue">


</body>
