---
title: "Import Spreadsheet Data Efficiently With R"
author: Timothy Deitz, Clinical Psychologist & Data Science Consultant
date: '2019-11-23'
categories:
  - Data Cleaning
tags:
  - Data Cleaning
output:
  blogdown::html_page:
    toc: no
---
<style>
body.blue { background-color:#659dbd;}
</style>

## Importing the Data

Real-world data is messy, irritating and cumbersome. Many tutorials on data wrangling with R simply don't prepare you for it. 

I begin by loading all required packages.

```{r, echo = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE, out.width = '100%')

```

```{r}
library(fs)
library(readr)
library(here)
library(purrr)
library(janitor)
library(dplyr)
library(stringr)
library(naniar)
library(plotly)
library(stringr)
library(forcats)

```

By using `here()` from the **fs** package (in contrast to an absolute file path), I ensure that people who have my data can run this script easily.

I then list the files present in my working directory. I want to import spreadsheets with a `.csv` extension and discard files with an `.xlsx` extension.

```{r}

fs::dir_ls(path = here())

```

I use a basic regular expression to capture only the `.csv` files. Then I import each one and store it in a list.

```{r}
csv_files<- fs::dir_ls(here(), regexp = "\\.csv$")

pain_data_raw<- csv_files %>% purrr::map( ~readr::read_csv(.x)) %>% 
  purrr::set_names("early_19", "mid_17", "mid_18")

```

Check that I now have a list of 3 dataframes.

```{r}

str(pain_data_raw, max.level = 0)

```

Yes!

I also want to check that each dataframe contains the same variables, and that those variables have identical data types (character, logical, numeric etc.). `compare_df_cols()` function from the **janitor** package lets you compare multiple dataframes according to the presence and data type of the variables they contain.

The output reveals a problem: some variables have a different data type depending on the dataframe they belong to. For example, the **EEnd_BPIPainInterference** column (which measures pain interference in a person's daily activities) shows up as a numeric variable for the first and third dataframes, but as a character varible for the second dataframe. This is most likely due to an error when data was entered into one of the spreadsheets.

```{r}

data_comparison<- janitor::compare_df_cols(pain_data_raw)

data_comparison %>% dplyr::filter(column_name == 'EEnd_BPIPainInterference')

```

When you have many columns (i.e. more than 100) and data entry mistakes are probable, it is sometimes best to manually import all columns as character variables, and then alter their type later. Therefore, I add `col_types = cols(.default = "c")` to the original code.

```{r}
pain_data_raw<- csv_files %>% purrr::map( ~readr::read_csv(.x, col_types = cols(.default = "c"))) %>% 
  purrr::set_names("early_19", "mid_17", "mid_18")

```

If I now re-examine the dataframe comparison, all variables are of the character type. But there is another problem. Certain variables appear in one dataframe and not the others, or appear in two of the dataframes but not the third. 

```{r}

head(data_comparison)

```

I **only** want to retain variables that are common to each dataframe. The first step in doing so is to store the names of the shared variables in a single character vector.

```{r}
matching_vars<- data_comparison %>% dplyr::filter_at(vars(-column_name), all_vars(!is.na(.))) %>% 
  dplyr::select(column_name) %>% dplyr::pull(1)

#Have a look

head(matching_vars)
```

I then iterate through each dataframe and select variables contained in that character vector.

```{r}

pain_data_complete<- pain_data_raw %>% purrr::map(~dplyr::select(.x, matching_vars))

```

Each dataframe should now have the same number of coluumns.

```{r}
purrr::map_dbl(pain_data_complete, ncol)
```

Success!!!

Finally, I join the three dataframes into a single dataframe.

```{r}
joined_data<- dplyr::bind_rows(pain_data_complete)

#Check it worked  

glimpse(joined_data) %>% print(n = 10)

```

Having imported and merged the data, I incorporate the data importation stesp into a reusable function for later use.

```{r}

import_data<- function() {

csv_files<- fs::dir_ls(here(), regexp = "\\.csv$")

pain_data_raw<- csv_files %>% purrr::map( ~readr::read_csv(.x, col_types = cols(.default = "c"))) %>% 
  purrr::set_names("early_19", "mid_17", "mid_18")

data_comparison<- janitor::compare_df_cols(pain_data_raw)

matching_vars<- data_comparison %>% dplyr::filter_at(vars(-column_name), all_vars(!is.na(.))) %>% 
  dplyr::select(column_name) %>% dplyr::pull(1)

pain_data_complete<- pain_data_raw %>% purrr::map(~dplyr::select(.x, matching_vars))

joined_data<- dplyr::bind_rows(pain_data_complete)

}

```

Now that I have the data, it's time to clean it up.

<body class = "blue">


</body>
