---
title: "Supercharged Data Cleaning with R"
author: Timothy Deitz, Clinical Psychologist & Data Science Consultant
date: '2019-11-23'
categories:
  - Data Cleaning
tags:
  - Data Cleaning
output:
  blogdown::html_page:
    toc: yes
---
<style>
body.blue { background-color:#2b3f60;}
</style>

# Supercharged Data Cleaning with R — Part One

## Importing the Data

Real-world data is messy, irritating and cumbersome. Many tutorials on data wrangling with R simply don't prepare you for it. 

I begin by loading all required packages.

```{r, echo = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE)

```

```{r}
library(fs)
library(readr)
library(here)
library(purrr)
library(janitor)
library(dplyr)
library(stringr)
library(naniar)
library(plotly)
library(stringr)
library(forcats)

```

The **fs** package is fantastic for anything that relates to file/folder navigation using R. By using **here()** not specifying an absolute file path (e.g. `tim_data\analysis\import`), other people who have the data can run my script straight away.

I next list all files in my working directory to locate the spreadsheets I need.

```{r}

fs::dir_ls(path = here())

```

Because I only want csv files, I use a basic regular expression to import only files that end in this extension. Then, I then read in each spreadsheet programatically, and the contents in a list of dataframes. 

```{r}
csv_files<- fs::dir_ls(here(), regexp = "\\.csv$")

pain_data_raw<- csv_files %>% purrr::map( ~readr::read_csv(.x)) %>% 
  purrr::set_names("early_19", "mid_17", "mid_18")

```

Check there is now a list of dataframes:

```{r}

str(pain_data_raw, max.level = 1)

```

The **readr** package automatically guesses the data type of each imported column. But it doesn't always guess right, which can create problems when you have large datasets. In theory, each of the three dataframes should have identical variables. But let's check this, using a handy function from the **janitor** package. This function lists variables in the far left column and shows whether that variable is present (and what data type it is) in each dataframe, the names of which are listed up top.

```{r}

data_comparison<- janitor::compare_df_cols(pain_data_raw)

data_comparison %>% dplyr::filter(column_name == 'EEnd_BPIPainInterference')

```

As you can see, the **EEnd_BPIPainInterference** column is specified as a numeric variable in two of the dataframes, but as a character varible in the other. To be clear, this variable (which measures how severely pain disrupts daily activities) should certainly be numeric. The character specification in one of the dataframes is probably caused by an errant fullstop or some other data entry error.

When you have lots of columns (> 100) and there are likely mistakes in data entry, it is often best to manually import all columns as character variables, and then alter their type as needed. I can do this by adding **col_types = cols(.default = "c")** to the original code.

```{r}
pain_data_raw<- csv_files %>% purrr::map( ~readr::read_csv(.x, col_types = cols(.default = "c"))) %>% 
  purrr::set_names("early_19", "mid_17", "mid_18")

```

If I now compare each dataframe again, all variables are of the character type. But there is another problem. Certain variables appear in one dataframe but not others, or appear in two dataframes but not in the third. 

```{r}

data_comparison

```

Since I want to analyse all dataframes combined, I opt to **only** retain variables that are common to each dataframe. To do this, I first store the names of those common variables in a character vector.

```{r}
matching_vars<- data_comparison %>% dplyr::filter_at(vars(-column_name), all_vars(!is.na(.))) %>% 
  dplyr::select(column_name) %>% dplyr::pull(1)

#Have a look

head(matching_vars)
```


I then use **map()** once again, to iterate through each dataframe and select only these variables

```{r}

pain_data_complete<- pain_data_raw %>% purrr::map(~dplyr::select(.x, matching_vars))

```

To be sure it worked, I can check that each dataframe now has the same number of columns. 

```{r}
purrr::map_dbl(pain_data_complete, ncol)
```

Success!!!

Having succeeded at importing and merging the data, it is a good idea to store the above steps into a reusable function.

```{r}

import_data<- function() {

csv_files<- fs::dir_ls(here(), regexp = "\\.csv$")

pain_data_raw<- csv_files %>% purrr::map( ~readr::read_csv(.x, col_types = cols(.default = "c"))) %>% 
  purrr::set_names("early_19", "mid_17", "mid_18")

data_comparison<- janitor::compare_df_cols(pain_data_raw)

matching_vars<- data_comparison %>% dplyr::filter_at(vars(-column_name), all_vars(!is.na(.))) %>% 
  dplyr::select(column_name) %>% dplyr::pull(1)

pain_data_complete<- pain_data_raw %>% purrr::map(~dplyr::select(.x, matching_vars))

joined_data<- dplyr::bind_rows(pain_data_complete)

}

```

Test that it works

```{r}

joined_data<- import_data()

head(glimpse(joined_data))

```


<body class = "blue">


</body>
