---
title: "Generating Clinical Cutoffs With R"
author: Timothy Deitz, Clinical Psychologist & Data Science Consultant
date: '2019-12-04'
categories:
  - Visualisation
tags:
  - Visualisation
output:
  blogdown::html_page:
    toc: no
---
<style>
body.blue { background-color:#659dbd;}
</style>



```{r, echo = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE, out.width = '100%')

```

```{r}
library(fs)
library(readr)
library(here)
library(purrr)
library(janitor)
library(dplyr)
library(stringr)
library(naniar)
library(plotly)
library(stringr)
library(forcats)
library(correlationfunnel)
library(magrittr)

```

In this post I use tidyverse data wrangling procedures to develop clinical cutoff criteria for psychological outcome measures. Then I use these criteria to examine the performance of a multidisciplinary pain management program for persistent pain.

First import the data:

```{r}

clean_data<- readRDS("clean_data.rds")

```

Next, for each outcome measure generate a function that will calculate change in scores on that measure, and whether or not that change met measure-specific benchmark criteria for a clinically significant improvement. I could have made a complex single function to accomplish this, but I prefer making several smaller ones.

To evalute program performance, it is critically important to exclude asymptomatic patients from analyses at the beginning. For example, if someone scored a 2 on the DASS-21 outcome measure, that person is asymptomatic for depression. They can't really "improve" further and so including them in the initial sample would water down the results!

```{r}
calc_sig_change_depression <- function(df) {
  df <-
    df %>% dplyr::mutate(
      change_depression = end_dass_depression - ref_dass_depression,
      sig_change_depression = dplyr::case_when(
        is.na(change_depression) |
          ref_dass_depression < 14 |
          ref_dass_depression_empty >= 2 |
          end_dass_depression_empty >= 2 ~ NA_character_,
        ref_dass_depression >= 14 &
          end_dass_depression < 14 & abs(change_depression) > 5 ~ "yes",
        ref_dass_depression >= 21 &
          end_dass_depression < 21 & abs(change_depression) > 5 ~ "yes",
        ref_dass_depression >= 28 &
          end_dass_depression < 28 & abs(change_depression) > 5 ~ "yes",
        TRUE ~ "no"
      )
    )
}


calc_sig_change_anxiety <- function(df) {
  df <-
    df %>% dplyr::mutate(
      change_anxiety = end_dass_anxiety - ref_dass_anxiety,
      sig_change_anxiety = dplyr::case_when(
        is.na(change_anxiety) |
          ref_dass_anxiety < 10 |
          ref_dass_anxiety_empty >= 2 |
          end_dass_anxiety_empty >= 2  ~ NA_character_,
        ref_dass_anxiety >= 10 &
          end_dass_anxiety < 10 & abs(change_anxiety) > 5 ~ "yes",
        ref_dass_anxiety >= 15 &
          end_dass_anxiety < 15 & abs(change_anxiety) > 5 ~ "yes",
        ref_dass_anxiety >= 20 &
          end_dass_anxiety < 20 & abs(change_anxiety) > 5 ~ "yes",
        TRUE ~ "no"
      )
    )
}




calc_sig_change_stress <- function(df) {
  df <-
    df %>% dplyr::mutate(
      change_stress = end_dass_stress - ref_dass_stress,
      sig_change_stress = dplyr::case_when(
        is.na(change_stress) |
          ref_dass_stress < 19 |
          ref_dass_stress_empty >= 2 |
          end_dass_stress_empty >= 2  ~ NA_character_,
        ref_dass_stress >= 19 &
          end_dass_stress < 19 & abs(change_stress) > 5 ~ "yes",
        ref_dass_stress >= 26 &
          end_dass_stress < 26 & abs(change_stress) > 5 ~ "yes",
        ref_dass_stress >= 34 &
          end_dass_stress < 34 & abs(change_stress) > 5 ~ "yes",
        TRUE ~ "no"
      )
    )
  
}



calc_sig_change_catastrophising <- function(df) {
  df <-
    df %>% dplyr::mutate(
      change_catastrophising = end_pcs_total - ref_pcs_total,
      sig_change_catastrophising = dplyr::case_when(
        is.na(change_catastrophising) |
          ref_pcs_total < 20 |
          ref_pcs_total_empty >= 2 |
          end_pcs_total_empty >= 2 ~ NA_character_,
        ref_pcs_total >= 20 &
          end_pcs_total < 20 & abs(change_catastrophising) > 5 ~ "yes",
        ref_pcs_total >= 30 &
          end_pcs_total < 30 & abs(change_catastrophising) > 5 ~ "yes",
        TRUE ~ "no"
      )
    )
  
}


calc_sig_change_self_efficacy <- function(df) {
  df <- df %>% dplyr::mutate(
    change_self_efficacy = end_pseq_total - ref_pseq_total,
    sig_change_self_efficacy = dplyr::case_when(
      is.na(change_self_efficacy) |
        ref_pseq_total > 30 |
        ref_pseq_empty >= 2 | end_pseq_empty >= 2  ~ NA_character_,
      ref_pseq_total <= 30 &
        end_pseq_total >= 31 & abs(change_self_efficacy) >= 7 ~ "yes",
      ref_pseq_total < 20 &
        end_pseq_total >= 20 & abs(change_self_efficacy) >= 7 ~ "yes",
      TRUE ~ "no"
    )
  )
}


calc_sig_change_pain_interference <- function(df) {
  df <- df %>% dplyr::mutate(
    change_pain_interference = end_bpi_pain_interference - ref_bpi_pain_interference,
    sig_change_pain_interference = dplyr::case_when(
      is.na(change_pain_interference) |
        ref_bpi_pain_interference < 5  ~ NA_character_,
      (
        abs(change_pain_interference) >= (ref_bpi_pain_interference * 0.3)
      ) & (end_bpi_pain_severity < ref_bpi_pain_severity) ~ "yes",
      TRUE ~ "no"
    )
  )
}



calc_sig_change_pain_severity <- function(df) {
  df <- df %>% dplyr::mutate(
    change_pain_severity = end_bpi_pain_severity - ref_bpi_pain_severity,
    sig_change_pain_severity = dplyr::case_when(
      is.na(change_pain_severity) |
        ref_bpi_pain_severity < 5  ~ NA_character_,
      (abs(change_pain_severity) >= (ref_bpi_pain_severity * 0.3)) &
        (end_bpi_pain_severity < ref_bpi_pain_severity) ~ "yes",
      TRUE ~ "no"
    )
  )
}

```

Now, apply each function to the dataframe to generate the required change and benchmark status variables.

```{r}

clean_data<- purrr::map_dfr(list(clean_data), ~{
  
  calc_sig_change_depression(.x) %>% 
  calc_sig_change_anxiety() %>% 
  calc_sig_change_stress() %>% 
  calc_sig_change_catastrophising() %>% 
  calc_sig_change_self_efficacy() %>% 
  calc_sig_change_pain_interference() %>% 
  calc_sig_change_pain_severity()
  
})

```

Gather the data, calculate program outcomes and thereby wrangle the data for plotting:

```{r}

plotting_data<- clean_data %>% dplyr::select( contains("sig_change"))

plotting_data_gathered<- plotting_data %>% tidyr::gather(key = "outcome", value = "sig_change_status")

plotting_data_summarised<- plotting_data_gathered %>% dplyr::group_by(outcome, sig_change_status) %>% dplyr::filter(!is.na(sig_change_status)) %>% 
  dplyr::summarise(n_cases = n()) %>% dplyr::ungroup() %>% dplyr::group_by(outcome) %>% 
  mutate(all_cases = round(sum(n_cases))) %>% dplyr::ungroup() %>% dplyr::mutate(pct_cases = round((n_cases / all_cases) * 100), outcome = stringr::str_replace(outcome, "sig_change_", ""), sig_change_status = fct_recode(sig_change_status, "Benchmark Improvement" = "yes", "No Benchmark Improvement" = "no"))
 
#It's tricky to arrange the bars when you have a stacked bar plot.The lines of code below allow us to do so

reordered_levels<- filter(plotting_data_summarised, sig_change_status == "Benchmark Improvement") %>% arrange(desc(pct_cases)) %>% dplyr::pull(outcome)


plotting_data_summarised<- plotting_data_summarised %>% dplyr::mutate(
  outcome = forcats::fct_relevel(outcome, reordered_levels)
   )

#Have a look

plotting_data_summarised

```

Finally, plot the outcomes:

```{r}
p<- ggplot(plotting_data_summarised, aes(x = outcome, y = pct_cases, fill = sig_change_status, 
                                     text = paste0("Number of Patients Showing ", sig_change_status, ": ", n_cases,  "<br>",
                                                  "Total Sample Size (Symptomatic Patients): ", all_cases))) + 
  geom_col() + geom_text(aes(label = paste0(pct_cases, "%")), position = position_stack(vjust = 0.5), size = 4) + 
  theme(panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.x = element_blank()
  ) + ylab("Percentage of Symptomatic Patients") + xlab("Outcome") +
  labs(fill = "Achieved Benchmark Improvement")  + 
  theme(legend.position = "bottom", legend.title = element_blank()) + coord_flip()


ggplotly(p, tooltip = "text") %>% layout(legend = list(orientation = "h", x = 0.4, y = -0.2))

```

Don't forget to hover over the plot to access info contained in the tooltip!

<body class = "blue">


</body>
