---
title: "Dynamic Exploratory Visualisation With R"
author: Timothy Deitz, Clinical Psychologist & Data Science Consultant
date: '2019-11-23'
categories:
  - Visualisation
tags:
  - Visualisation
output:
  blogdown::html_page:
    toc: no
---
<style>
body.blue { background-color:#659dbd;}
</style>

```{r, echo = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE, out.width = '100%')

```


```{r}
library(fs)
library(readr)
library(here)
library(purrr)
library(janitor)
library(dplyr)
library(stringr)
library(naniar)
library(plotly)
library(stringr)
library(forcats)
library(magrittr)
```

In previous articles, I imported and cleaned a dataset of chronic pain outcomes for an allied health treatment program.

I start by reading in this cleaned data.

```{r}

clean_data<- readRDS("clean_data.rds")

```

I select variables of interest to focus on in analyses.

```{r}
vars_of_interest<-  dplyr::select(clean_data, patient_id, sex, ref_hsugp, ref_hsu_allied_health, ref_hsu_specialist, ref_hsu_hospital, ref_hsued, ref_unemployed_pain, ref_daily_morphine, ref_bpi_work, ref_hsu_diagnostic_tests, ref_opioid_freq, ref_full_time, ref_bpi_relations, ref_pcs_total, ref_pseq_total, ref_pcs_helplessness, ref_pcs_magnification,
ref_pcs_rumination, ref_bpi_sleep, ref_bpi_average, ref_weight, ref_dass_depression,
ref_dass_anxiety, ref_dass_stress)
```

Next, I use some tidy evaluation and functional programing principles write functions that dynamically iterate through these variables and create exploratory plots.

I begin with boxpots 

```{r}
show_all_boxplots<- function(df, label) {
  
  identifier<- enquo(label)
  
  #Iterate over all numeric variables
  #Iterate over names of these variables (needed for labels)
  
  num_vars<- dplyr::select_if(df, is.numeric)
  num_var_names<- names(num_vars)
  
  plot_list<- purrr::pmap(list(num_vars, num_var_names, dplyr::select(df, !!identifier)), ~{
    
    
    plot<- ggplot(df, aes(x = "", y = ..1, 
                                    text = paste("Identifier:", ..3, "<br>",
                                                 ..2, "value:", ..1, "<br>",
                                                 ..2, "mean:", round(mean(..1, na.rm = TRUE), 2), "<br>",
                                                 ..2, "sd:", round(sd(..1, na.rm = TRUE), 2), "<br>",
                                                 ..2, "N outliers:", 
                                                 nrow(dplyr::filter(df, 
                                                                    ..1 > (mean(..1, na.rm = TRUE) + 2 * sd(..1, na.rm = TRUE))))
                                                 
                                    ))) +
      
      # make boxplot transparent with alpha = 0
      geom_boxplot(alpha = 0) +
      
      # If a given point is an outlier, colour it red
      
      geom_jitter(color = ifelse(..1 > (mean(..1, na.rm = TRUE) + 
                                         2 * sd(..1, na.rm = TRUE)), 'red', 'steelblue'), alpha = 0.3) +
      
      theme(axis.title.x=element_blank()) + ylab(paste(..2))
    
    ggplotly(plot, tooltip = "text")
    
    
  })
  
  #Due to plotly bug, we need to set opacity of boxplot markers
  #to zero, so that outliers aren't duplicated on the plot.
  #Below, go inside each plot object in the list to do this.
  
  for(i in 1:length(plot_list)) {
    
    plot_list[[i]][["x"]][["data"]][1][[1]][["marker"]][["opacity"]]<- 0
    
  }
  
  plot_list<- plot_list %>% purrr::set_names(num_var_names)
  
  
  return(plot_list)
}

```


Test the function:

```{r}
box_list<- show_all_boxplots(vars_of_interest, sex)

#Number of visits to a specialist in the past 3 months 

box_list[["ref_hsu_specialist"]]

```

Next I write a function to dynamically plot histograms

```{r}
show_all_histograms<- function(df) {
  
  numeric_data<- dplyr::select_if(df, is.numeric)
  
  plot_list<- purrr::map2(numeric_data, names(numeric_data), ~{
    
    plot<- ggplot(numeric_data, aes(x = .x)) + geom_histogram() + xlab(paste0(.y)) + ylab(paste0(.y))
    
    ggplotly(plot)
    
  })
  
  plot_list<- plot_list %>% purrr::set_names(names(numeric_data))
  
}

```

Test the function:

```{r}

hist_list<- show_all_histograms(vars_of_interest)

#Plot DASS-21 stress for all participants

hist_list[["ref_dass_stress"]]

```


Now write a function for density plots

```{r}
show_all_density<- function(df, gaussian_kernel = 2.5) {
  
  numeric_data<- dplyr::select_if(df, is.numeric)
  
  plot_list<- purrr::map2(numeric_data, names(numeric_data), ~ {
    
    #Show the skew and kurtosis of each plot in a tooltip
    
    density_plots<- ggplot(numeric_data, aes(x = .x,
                                             text = paste("skew:", round(psych::skew(.x), 2), 
                                                          "<br>",
                                                          "kurtosis:", round(psych::kurtosi(.x), 2)))) +
      
      geom_density(fill = 'steelblue', bw = gaussian_kernel,  
                   alpha = 0.7) +
      
      #Add the rug to the bottom of the density plot
      
      geom_rug() + xlab(paste0(.y))
    
    plotly::ggplotly(density_plots)
    
  })
  
  plot_list<- plot_list %>% purrr::set_names(names(numeric_data))
  
}

```

Test the function:

```{r}
density_list<- show_all_density(vars_of_interest)

#Plot pain-related sleep disturbance

density_list[["ref_bpi_sleep"]]

```

Write a function to make QQ plots

```{r}

show_all_qqplots<- function(df) {
  
  numeric_data<- dplyr::select_if(df, is.numeric)
  
  plot_list<- purrr::map(numeric_data, ~{
    
    qqplots<- ggplot(numeric_data, aes(sample = .x)) + stat_qq() + stat_qq_line()
    
    plotly::ggplotly(qqplots)
    
  })
  
  plot_list<- plot_list %>% purrr::set_names(names(numeric_data))
  
}



```

Test the function:

```{r}

qq_list<- show_all_qqplots(vars_of_interest)

#Plot visits to a GP in the past 3 months

qq_list[["ref_hsugp"]]

```

Combine all of the above plot types into a subplot, accessible by variable name:

```{r}
show_distributions<- function(df, label, gaussian_kernel = 2.5) {
  
  label_unquoted<- rlang::enquo(label)

  numeric_data<- dplyr::select_if(df, is.numeric)
  
  #Make the list of boxplots
  
  boxplot_list<- show_all_boxplots(df, !!label_unquoted)
  
  #Make the list of density plots
  
  density_list<- show_all_density(df, gaussian_kernel)
  
  #Make the list of histograms
  
  histogram_list<- show_all_histograms(df)
  
  #Make the list of QQ plots
  
  qq_list<- show_all_qqplots(df)
  
  #Iterate through the plot-specific lists to make a new list 
 
  plots_list<- purrr::pmap(list(boxplot_list, density_list, histogram_list, qq_list),
                           ~plotly::subplot(..1, ..2, ..3, ..4, nrows = 2, titleY = TRUE))
  
  plots_list<- plots_list %>% purrr::set_names(names(numeric_data))
  
}  

```

Test the function:

```{r}
dist_list<- show_distributions(vars_of_interest, sex)

#Plot average pain intensity

dist_list[["ref_bpi_average"]]

#Plot the impact of pain on relationships

dist_list[["ref_bpi_sleep"]]

```



<body class = "blue">


</body>

