---
title: "Create Interactive Scatterplots Programmatically With R"
author: Timothy Deitz, Clinical Psychologist & Data Science Consultant
date: '2019-11-24'
categories:
  - Visualisation
tags:
  - Visualisation
output:
  blogdown::html_page:
    toc: no
---
<style>
body.blue { background-color:#659dbd;}
</style>



```{r, echo = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE, out.width = '100%')

```

```{r}
library(fs)
library(readr)
library(here)
library(purrr)
library(janitor)
library(dplyr)
library(stringr)
library(naniar)
library(plotly)
library(stringr)
library(forcats)
library(correlationfunnel)
library(magrittr)

```

In this article, I programatically generate interactive scatterplots using the persistent pain dataset that I imported and cleaned in earlier posts. I want to explore relationships between features to guide subsequent modelling. 

First import the data:

```{r}

clean_data<- readRDS("clean_data.rds")

```

Then make things more manageable by reducing the dataset down to variables of interest for exploratory analyses.

```{r}
vars_of_interest<-  dplyr::select(clean_data, patient_id, sex, ref_hsugp, ref_hsu_allied_health, ref_hsu_specialist, ref_hsu_hospital, ref_hsued, ref_unemployed_pain, ref_daily_morphine, ref_bpi_work, ref_hsu_diagnostic_tests, ref_opioid_freq, ref_full_time, ref_bpi_relations, ref_pcs_total, ref_pseq_total, ref_pcs_helplessness, ref_pcs_magnification, ref_pcs_rumination, ref_bpi_sleep, ref_bpi_average, ref_bpi_pain_severity, ref_weight, ref_dass_depression,
ref_dass_anxiety, ref_dass_stress, pain_duration)
```

I am particularly interested in the outcomes of sleep disturbance and pain self-efficacy (confidence to perform a range of daily activities despite pain). So I write my own function based upon functions within the **corrr** package to quickly identify the correlates of my sleep disturbance and pain self-efficacy variables. 

```{r}

find_correlations<- function(df) {
  
  numeric_data<- dplyr::select_if(df, is.numeric)
  
  numeric_var_names<- rlang::syms(names(numeric_data))
  
  correlations<- corrr::correlate(numeric_data)
  
  correlation_by_var_list<- purrr::map(numeric_var_names, ~{corrr::focus(correlations, !!.x) %>% 
      dplyr::filter(!!.x >= 0.3 | !!.x <= -0.3) %>% dplyr::arrange(desc(!!.x))})
  
  corelation_by_var_list<- correlation_by_var_list %>% purrr::set_names(names(numeric_data))
}

```

Call `find_correlations()` to generate a list of of correlations greater than 0.3 for all my variables of interest.

```{r}
#Make the list of  correlations > 0.3 by variable.

corr_by_var<- find_correlations(vars_of_interest)
```

Subset the list to pluck out my interesting variables.

First, sleep disturbance:

```{r}
corr_by_var[["ref_bpi_sleep"]]
```

The highest raw correlation tells us that people who report greater sleep disturbance also tend to report worse relationship dynamics due to pain. Makes sense.

Now I look at the correlates of pain self-efficacy

```{r}
corr_by_var[["ref_pseq_total"]]
```

Curiously, the high raw correlation suggests that people whose relationships are more impacted by pain tend to have lower confidence to perform daily activities in the presence of pain!

I can also use the **correlationfunnel** package to help with identifying features for modelling. First, I need to impute missing values, which I can do easily with **naniar**.

```{r}
imputed_vars<- vars_of_interest %>% dplyr::select(-patient_id) %>% naniar::impute_median_if(is.numeric) 
```

Then I create a feature set of binary variables:

```{r}
binarized_vars<- imputed_vars %>% correlationfunnel::binarize(n_bins = 5, thresh_infreq = 0.01, name_infreq = "OTHER", one_hot = TRUE)

#Have a look

dplyr::glimpse(binarized_vars)

```

Finally, I create a correlation funnel plot to show which features are most correlated with the target outcome (pain-self-efficacy).

```{r}
binarized_vars %>% correlationfunnel::correlate(ref_pseq_total__35_Inf) %>% correlationfunnel::plot_correlation_funnel()
```

Features at the top of the plot are the most related to pain self-efficacy. We can see by the right outward curve, patients with lower levels of depression, relationship disruption (due to pain), stress, hopelessness, sleep disturbance (and a few related constructs) have greater confidence to perform daily activities in the presence of pain! People who have only had pain 3-5 months also have higher pain self-efficacy than others. 

To make sure that we are not being misled by raw numbers, it is important to plot the variables of interest.`ggpairs()` from the **GGally** package creates a matrix of scatterplots, based on the columns that are passed to it. Unfortunately, this function is not very helpful if you have many numeric variables and your computer may even in the process.

With large datasets, I think it can be helpful to create a named list of scatterplots, the elements of which can be scrolled through (using the R Studio viewer for example) or selected based on the name of the variable in question. 

Here is the code required to generate such a function. Explanatory comments are embedded.

```{r}

show_all_scatterplots<- function(df) {
  
  df<- dplyr::select_if(df, is.numeric)
  
  #Create a two-column dataframe with all possible pairs of numeric variable names
  
  numeric_pairs<- tibble::tibble(x_variable = names(df), 
                                 y_variable = x_variable) %>% expand.grid() 
  
  #Iterate over the two columns of numeric variable names we've created and use them 
  #to select variable from mtcars programatically
  
  plot_list<- purrr::map2(numeric_pairs$x_variable, numeric_pairs$y_variable, ~ { 
    
    #Create linear models for each pair of variables — required for fitting a line 
    #to each plot
    
    model<- lm(df[[.y]] ~ df[[.x]], data = df)
    
    #Create the plotly base for the plot. Programmatic mapping in plotly requires us 
    #to specify variables using square bracket notation, rather than the 
    #usual x = ~x, y = ~y. I don't know why.
    
    plotting_df<- tibble::tibble(x_plot_var = df[[.x]], y_plot_var = df[[.y]]) %>% dplyr::filter(complete.cases(.)) 
    
    plot_ly(plotting_df, x = ~x_plot_var, y = ~y_plot_var) %>%
      add_markers(x = ~jitter(x_plot_var), y = ~jitter(y_plot_var), opacity = 0.5, showlegend = FALSE) %>%
      
      #Add the line of best fit to each plot
      
      add_lines(y = ~fitted(model))  %>% 
      layout(title = paste0(.x, " vs ", .y,":", " 
                            Correlation = ", round(cor(df[[.x]],
                                                       df[[.y]], use = "pairwise.complete.obs"), 2)),
                                            xaxis = list(title = paste(.x)),
                                            yaxis = list(title = paste(.y))) %>% config(displayModeBar = F)
  })
  
  plot_names<- numeric_pairs %>% tidyr::unite("plot_labels", x_variable:y_variable, sep = "AND", remove = TRUE) %>% dplyr::pull(plot_labels)
  
  plot_list<- plot_list %>% purrr::set_names(plot_names)
  
}
```

Following from our correlation analysis, let's look at the relationship between pain self-efficacy pain severity.

```{r}

scatter_by_var<- show_all_scatterplots(vars_of_interest)

scatter_by_var[["ref_bpi_pain_severityANDref_pseq_total"]]

```

The scatterplot is informative, but it would be good to take subgroups into account. Doing so is important to avoid Simpson's Paradox!

The code for dynamically creating grouped scatterplots is shown below:

```{r}
show_all_scatterplots_grouped<- function(df, point_identifier,
                                         grouping_variable) {
  
  group<- enquo(grouping_variable)
  
  label<- enquo(point_identifier)
  
  numeric_pairs<- tibble::tibble(x_variable = names(dplyr::select_if(df, 
                                                                     is.numeric)),
                                 y_variable = x_variable) %>% expand.grid()
  
  
  legend_title <- list(yref='paper',xref="paper",y=1.05,x=1.2, 
                       text= paste(dplyr::select(df, !!group) %>% names()), showarrow=F)

  
  
  plot_list<- purrr::map2(numeric_pairs$x_variable, numeric_pairs$y_variable, ~ {
    
    #Store the name of the x-axis & y-axis variables 
    #(we won't be able to acces .x or .y once we start a new map() statement)
    
    x_var<- .x   
    
    y_var<- .y
    
    #Make a list of dataframes, based on levels of grouping variable 
    #selected by the user
    
    splitting_group<- dplyr::select(df, !!group) %>% pull(1)
    
    split_df<- split(df, splitting_group)
    
    #For each dataframe in the list, add a group-specific correlation value 
    #and group size
    
    merged_df<- purrr::map(split_df, 
                           ~dplyr::mutate(., 
                                          correlation = round(cor(.x[[paste(x_var)]], .x[[paste(y_var)]], 
                                                                  use = "pairwise.complete.obs"), 2),
                                          N = nrow(.x)
                           ))
    
    #Merge the list of dataframes back into a single dataframe
    
    merged_df<- do.call(rbind, merged_df)
    
    lm_grouping_var<- dplyr::select(merged_df, !!group) %>% pull()
    
    #Create a linear model to be able to plot the line of best fit
    #We specify grouped linear models using 
    #multiplication notation: y ~ x * categorical_variable
    
    model<- lm(merged_df[[paste(.y)]] ~ merged_df[[paste(.x)]]*as.factor(lm_grouping_var), 
               data = merged_df)
    
    #Make the plotly base and customise the hover info
    
    plotly_identifier_var<- dplyr::select(merged_df, !!label) %>% pull()
    
    plotting_df<- tibble::tibble(plotly_x_var = merged_df[[paste(.x)]], plotly_y_var = merged_df[[paste(.y)]], correlation = merged_df[["correlation"]], N = merged_df[["N"]],
                   plotly_grouping_var = lm_grouping_var, plotly_identifier_var = plotly_identifier_var) %>% dplyr::filter(complete.cases(plotly_x_var, plotly_y_var, plotly_identifier_var, plotly_grouping_var)) 
    
    plot_ly(plotting_df, x = ~plotly_x_var, y = ~plotly_y_var, 
            color = ~as.factor(plotly_grouping_var),
            hoverinfo = "text", 
            text = ~paste("Identifier:", plotly_identifier_var, "<br>",
                          .x, ": ", plotly_x_var, "<br>",
                          .y, ": ", plotly_y_var, "<br>",
                          "Group Correlation:", correlation, "<br>",
                          "Group N:", N
                          
            )) %>%
      
      add_markers(x = ~jitter(plotly_x_var), y = ~jitter(plotly_y_var), opacity = 0.5, showlegend = TRUE) %>%
      
      add_lines(y = ~fitted(model), showlegend = FALSE) %>%
      
      
      layout(title = paste0(.x, " vs ", .y,":", " Correlation = ", 
                            round(cor(merged_df[[paste(.x)]], merged_df[[paste(.y)]], use = "pairwise.complete.obs"), 2)),
             xaxis = list(title = paste(.x)),
             yaxis = list(title = paste(.y)), annotations = legend_title) %>% config(displayModeBar = F)
    
  })
  
  plot_names<- numeric_pairs %>% tidyr::unite("plot_labels", x_variable:y_variable, sep = "AND", remove = TRUE) %>% dplyr::pull(plot_labels)
  
  plot_list<- plot_list %>% purrr::set_names(plot_names)
  
}


```

Let's look at the relationship between pain self-efficacy and pain severity based on pain duration.

```{r}
grouped_scatterplots<- show_all_scatterplots_grouped(vars_of_interest, point_identifier = patient_id, grouping_variable = pain_duration)

grouped_scatterplots[["ref_pseq_totalANDref_bpi_pain_severity"]]

```

Look at how the relationship between pain severity and confidence to perform activities almost disappears for people who have only had pain a few months. But we also notice that this is based on a small sample size of only 8 patients — preventing any definitive conclusions.

<body class = "blue">


</body>
