---
title: "Create Interactive Scatterplots Programmatically With R"
author: Timothy Deitz, Clinical Psychologist & Data Science Consultant
date: '2019-11-24'
categories:
  - Visualisation
tags:
  - Visualisation
output:
  blogdown::html_page:
    toc: no
---
<style>
body.blue { background-color:#659dbd;}
</style>



```{r, echo = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE, out.width = '100%')

```

```{r}
library(fs)
library(readr)
library(here)
library(purrr)
library(janitor)
library(dplyr)
library(stringr)
library(naniar)
library(plotly)
library(stringr)
library(forcats)

```

In this article, I programatically generate interactive scatterplots using the persistent pain dataset that I imported and cleaned in earlier posts. 


```{r}

clean_data<- readRDS("clean_data.rds")

```

I select variables of interest for exploratory analyses.

```{r}
vars_of_interest<-  dplyr::select(clean_data, patient_id, sex, ref_hsugp, ref_hsu_allied_health, ref_hsu_specialist, ref_hsu_hospital, ref_hsued, ref_unemployed_pain, ref_daily_morphine, ref_bpi_work, ref_hsu_diagnostic_tests, ref_opioid_freq, ref_full_time, ref_bpi_relations, ref_pcs_total, ref_pseq_total, ref_pcs_helplessness, ref_pcs_magnification,
ref_pcs_rumination, ref_bpi_sleep, ref_bpi_average, ref_weight, ref_dass_depression,
ref_dass_anxiety, ref_dass_stress, pain_duration)
```

The `ggpairs()` function from **GGally** creates a matrix of scatterplots, based on the columns that are passed to it. But the function is not very helpful if you have a large number of numeric variables and your computer will probably crash in the process.

With large datasets, I think it can be helpful to create a named list of scatterplots, the elements of which can be scrolled through (using the R Studio viewer for example) or selected based on the name of the variable in question. 

Here is the code required to generate such a function. Explanatory comments are embedded.

```{r}

show_all_scatterplots<- function(df) {
  
  df<- dplyr::select_if(df, is.numeric)
  
  #Create a two-column dataframe with all possible pairs of numeric variable names
  
  numeric_pairs<- tibble::tibble(x_variable = names(df), 
                                 y_variable = x_variable) %>% expand.grid() 
  
  #Iterate over the two columns of numeric variable names we've created and use them 
  #to select variable from mtcars programatically
  
  plot_list<- purrr::map2(numeric_pairs$x_variable, numeric_pairs$y_variable, ~ { 
    
    #Create linear models for each pair of variables — required for fitting a line 
    #to each plot
    
    model<- lm(df[[.y]] ~ df[[.x]], data = df)
    
    #Create the plotly base for the plot. Programmatic mapping in plotly requires us 
    #to specify variables using square bracket notation, rather than the 
    #usual x = ~x, y = ~y. I don't know why.
    
    plotting_df<- tibble::tibble(x_plot_var = df[[.x]], y_plot_var = df[[.y]]) %>% dplyr::filter(complete.cases(.)) 
    
    plot_ly(plotting_df, x = ~x_plot_var, y = ~y_plot_var) %>%
      add_markers(x = ~jitter(x_plot_var), y = ~jitter(y_plot_var), opacity = 0.5, showlegend = FALSE) %>%
      
      #Add the line of best fit to each plot
      
      add_lines(y = ~fitted(model))  %>% 
      layout(title = paste0(.x, " vs ", .y,":", " 
                            Correlation = ", round(cor(df[[.x]],
                                                       df[[.y]], use = "pairwise.complete.obs"), 2)),
                                            xaxis = list(title = paste(.x)),
                                            yaxis = list(title = paste(.y))) %>% config(displayModeBar = F)
  })
  
  plot_names<- numeric_pairs %>% tidyr::unite("plot_labels", x_variable:y_variable, sep = "AND", remove = TRUE) %>% dplyr::pull(plot_labels)
  
  plot_list<- plot_list %>% purrr::set_names(plot_names)
  
}
```

Look at the relationship between depression and chronic pain-related relationship problems

```{r}

scatter_by_var<- show_all_scatterplots(vars_of_interest)

scatter_by_var[["ref_bpi_relationsANDref_dass_depression"]]

```

Sometimes, we also want to look at subgroups within scatterplots. Doing so is important to avoid Simpson's Paradox!

The code for dynamically creating grouped scatterplots is shown below:

```{r}
show_all_scatterplots_grouped<- function(df, point_identifier,
                                         grouping_variable) {
  
  group<- enquo(grouping_variable)
  
  label<- enquo(point_identifier)
  
  numeric_pairs<- tibble::tibble(x_variable = names(dplyr::select_if(df, 
                                                                     is.numeric)),
                                 y_variable = x_variable) %>% expand.grid()
  
  
  legend_title <- list(yref='paper',xref="paper",y=1.05,x=1.2, 
                       text= paste(dplyr::select(df, !!group) %>% names()), showarrow=F)

  
  
  plot_list<- purrr::map2(numeric_pairs$x_variable, numeric_pairs$y_variable, ~ {
    
    #Store the name of the x-axis & y-axis variables 
    #(we won't be able to acces .x or .y once we start a new map() statement)
    
    x_var<- .x   
    
    y_var<- .y
    
    #Make a list of dataframes, based on levels of grouping variable 
    #selected by the user
    
    splitting_group<- dplyr::select(df, !!group) %>% pull(1)
    
    split_df<- split(df, splitting_group)
    
    #For each dataframe in the list, add a group-specific correlation value 
    #and group size
    
    merged_df<- purrr::map(split_df, 
                           ~dplyr::mutate(., 
                                          correlation = round(cor(.x[[paste(x_var)]], .x[[paste(y_var)]], 
                                                                  use = "pairwise.complete.obs"), 2),
                                          N = nrow(.x)
                           ))
    
    #Merge the list of dataframes back into a single dataframe
    
    merged_df<- do.call(rbind, merged_df)
    
    lm_grouping_var<- dplyr::select(merged_df, !!group) %>% pull()
    
    #Create a linear model to be able to plot the line of best fit
    #We specify grouped linear models using 
    #multiplication notation: y ~ x * categorical_variable
    
    model<- lm(merged_df[[paste(.y)]] ~ merged_df[[paste(.x)]]*as.factor(lm_grouping_var), 
               data = merged_df)
    
    #Make the plotly base and customise the hover info
    
    plotly_identifier_var<- dplyr::select(merged_df, !!label) %>% pull()
    
    plotting_df<- tibble::tibble(plotly_x_var = merged_df[[paste(.x)]], plotly_y_var = merged_df[[paste(.y)]], correlation = merged_df[["correlation"]], N = merged_df[["N"]],
                   plotly_grouping_var = lm_grouping_var, plotly_identifier_var = plotly_identifier_var) %>% dplyr::filter(complete.cases(plotly_x_var, plotly_y_var, plotly_identifier_var, plotly_grouping_var)) 
    
    plot_ly(plotting_df, x = ~plotly_x_var, y = ~plotly_y_var, 
            color = ~as.factor(plotly_grouping_var),
            hoverinfo = "text", 
            text = ~paste("Identifier:", plotly_identifier_var, "<br>",
                          .x, ": ", plotly_x_var, "<br>",
                          .y, ": ", plotly_y_var, "<br>",
                          "Group Correlation:", correlation, "<br>",
                          "Group N:", N
                          
            )) %>%
      
      add_markers(x = ~jitter(plotly_x_var), y = ~jitter(plotly_y_var), opacity = 0.5, showlegend = TRUE) %>%
      
      add_lines(y = ~fitted(model), showlegend = FALSE) %>%
      
      
      layout(title = paste0(.x, " vs ", .y,":", " Correlation = ", 
                            round(cor(merged_df[[paste(.x)]], merged_df[[paste(.y)]], use = "pairwise.complete.obs"), 2)),
             xaxis = list(title = paste(.x)),
             yaxis = list(title = paste(.y)), annotations = legend_title) %>% config(displayModeBar = F)
    
  })
  
  plot_names<- numeric_pairs %>% tidyr::unite("plot_labels", x_variable:y_variable, sep = "AND", remove = TRUE) %>% dplyr::pull(plot_labels)
  
  plot_list<- plot_list %>% purrr::set_names(plot_names)
  
}


```

Look at the relationship between pain intensity and stress, based on the duration of chronic pain

```{r}
grouped_scatterplots<- show_all_scatterplots_grouped(vars_of_interest, point_identifier = patient_id, grouping_variable = pain_duration)

grouped_scatterplots[["ref_dass_stressANDref_bpi_average"]]

```

Hope this is helpful, feel free to reach out and let me know your thoughts.


<body class = "blue">


</body>
